---
title: "ARIMA/SARIMA"
author: "Hritik"
date: "2025-03-03"
output: html_document
---

```{r}

# Load required libraries
library(tidyverse)
library(lubridate)
library(forecast)
library(tseries)
library(readr)
library(dplyr)
library(ggplot2)
library(lmtest)
library(FinTS)

# Load dataset
setwd("/Users/hritikjhaveri/Desktop/UChicago/Q2/Time_Series_Analysis_Forcasting/Final_Project")
getwd()

data <- read.csv("London_AQI_Daily.csv")

# Convert 'Date' column to datetime format
data$Date <- as.Date(data$Date)

df <- data.frame(Date = data$Date, USAQI = data$USAQI)

# View first few rows
head(df)

# Check data types
str(df)

# Check for missing values
sum(is.na(df))

# Summary stats for AQI
summary(df$USAQI)

```

# 2. Visualize Time Series

```{r}

# Line plot
ggplot(df, aes(x = Date, y = USAQI)) +
  geom_line(color = "blue") +
  labs(title = "AQI in London from 2013-2024", x = "Year", y = "AQI in London") +
  theme_minimal()

```

# 3. Loading data into TS object

```{r}

# Decompose Time Series
ts_data <- ts(df$USAQI)

str(ts_data)

```

# 4. ADF Test

```{r}

# Augmented Dickey-Fuller (ADF) test
adf.test(ts_data)

```

# 5. KPSS Test

```{r}

# Kwiatkowski-Phillips-Schmidt-Shin (KPSS) test
kpss.test(ts_data)

```

# 6. Autocorrelation Function

```{r}

# ACF plots
acf(ts_data)

```

# Observations from ACF: The ACF plot shows a slow decay, which is a sign of non-stationarity. There are no clear seasonal spikes, so it does not strongly indicate seasonality. The gradual decline suggests an AR or ARMA process rather than pure MA.

# 7. Partial Autocorrelation Function

```{r}

# PACF plots
pacf(ts_data)

```

#Observations from PACF: The PACF cuts off sharply after lag 1, meaning the series may follow an AR(1) process. There are no further significant spikes after lag 1, which supports an AR(1) model.

```{r}

# First order difference
ts_data_diff <- diff(ts_data)

print(adf.test(ts_data_diff))
print(kpss.test(ts_data_diff))

```

```{r}

acf(ts_data_diff)

```

```{r}

pacf(ts_data_diff)

```

**p = 1 or 2, d = 1, q = 1**

# 8. Splitting data into training and testing sets

```{r}

# Find train indices
train_start_index <- which(df$Date == as.Date("2013-01-02"))
train_end_index <- which(df$Date == as.Date("2024-12-31"))

# Extract train window using indices
diff_train_data <- window(ts_data_diff, start = train_start_index, end = train_end_index)

# Find test indices
test_start_index <- which(df$Date == as.Date("2025-01-01"))
test_end_index <- which(df$Date == as.Date("2025-02-28"))

# Extract test window using indices
diff_test_data <- window(ts_data_diff, start = test_start_index, end = test_end_index)

```

```{r}

# Extract train window using indices
train_data <- window(ts_data, start = train_start_index, end = train_end_index)

# Extract test window using indices
test_data <- window(ts_data, start = test_start_index, end = test_end_index)

```


# 9.0 AUTO.ARIMA Model 

```{r}

arima_model <- auto.arima(train_data)

summary(arima_model)

```

```{r}

# Forecast for test period
forecast_values <- forecast(arima_model, h = length(test_data))

# Plot actual vs forecast
autoplot(forecast_values) + 
  autolayer(test_data, series="Actual") + 
  ggtitle("ARIMA Forecast vs Actual Data") + 
  theme_minimal()

```

```{r}

# Ljung-Box Test for residual autocorrelation
Box.test(arima_model$residuals, lag = 20, type = "Ljung-Box")

```

```{r}

# Normality Tests
shapiro_test <- shapiro.test(arima_model$residuals)
print(shapiro_test)

jarque_bera_test <- jarque.bera.test(arima_model$residuals)
print(jarque_bera_test)

checkresiduals(arima_model)

```

# 9.1 ARIMA Model (1,1,1)

```{r}

# Fit ARIMA model
arima_model_1 <- Arima(diff_train_data, order = c(1, 1, 1), seasonal = FALSE)

# Summary of ARIMA model
summary(arima_model_1)

```

```{r}

# Forecast for test period
forecast_values <- forecast(arima_model_1, h = length(diff_test_data))

# Plot actual vs forecast
autoplot(forecast_values) + 
  autolayer(diff_test_data, series="Actual") + 
  ggtitle("ARIMA Forecast vs Actual Data") + 
  theme_minimal()

```

```{r}

# Ljung-Box Test for residual autocorrelation
Box.test(arima_model_1$residuals, lag = 20, type = "Ljung-Box")

```

```{r}

# Normality Tests
shapiro_test <- shapiro.test(arima_model_1$residuals)
print(shapiro_test)

jarque_bera_test <- jarque.bera.test(arima_model_1$residuals)
print(jarque_bera_test)

checkresiduals(arima_model_1)

```

# 9.2 ARIMA Model (2,1,1)

```{r}

# Fit ARIMA model
arima_model_2 <- Arima(diff_train_data, order = c(2, 1, 1), seasonal = FALSE)

# Summary of ARIMA model
summary(arima_model_2)

```


```{r}

# Forecast for test period
forecast_values <- forecast(arima_model_2, h = length(diff_test_data))

# Plot actual vs forecast
autoplot(forecast_values) + 
  autolayer(diff_test_data, series="Actual") + 
  ggtitle("ARIMA Forecast vs Actual Data") + 
  theme_minimal()

```

```{r}

# Ljung-Box Test for residual autocorrelation
Box.test(arima_model_2$residuals, lag = 20, type = "Ljung-Box")

```

```{r}

# Normality Tests
shapiro_test <- shapiro.test(arima_model_2$residuals)
print(shapiro_test)

jarque_bera_test <- jarque.bera.test(arima_model_2$residuals)
print(jarque_bera_test)

checkresiduals(arima_model_2)

```

# 10.0 SARIMA Model

```{r}

# Fit SARIMA model
sarima_model <- auto.arima(train_data, seasonal = TRUE)

# Summary of SARIMA model
summary(sarima_model)

```

```{r}

# Forecast for test period
forecast_values <- forecast(sarima_model, h = length(test_data))

# Plot actual vs forecast
autoplot(forecast_values) + 
  autolayer(test_data, series="Actual") + 
  ggtitle("ARIMA Forecast vs Actual Data") + 
  theme_minimal()

```

```{r}

# Ljung-Box Test for residual autocorrelation
Box.test(sarima_model$residuals, lag = 20, type = "Ljung-Box")

```

```{r}

# Normality Tests
shapiro_test <- shapiro.test(sarima_model$residuals)
print(shapiro_test)

jarque_bera_test <- jarque.bera.test(sarima_model$residuals)
print(jarque_bera_test)

checkresiduals(sarima_model)

```
# 11

```{r}

# RMSE
print('Arima')
accuracy(arima_model)
print('Arima_diff (1,1,1)')
accuracy(arima_model_1)
print('Arima_diff (2,1,1)')
accuracy(arima_model_2)
print('Sarima')
accuracy(sarima_model)

```

```{r}

# Perform ARCH test

arch_test <- ArchTest(arima_model$residuals, lags = 20)
print(arch_test)

arch_test_1 <- ArchTest(arima_model_1$residuals, lags = 20)
print(arch_test)

arch_test_2 <- ArchTest(arima_model_1$residuals, lags = 20)
print(arch_test)

arch_test_3 <- ArchTest(sarima_model$residuals, lags = 20)
print(arch_test)

```

# 12.0 AUTO.ARIMA Model log

```{r}

log_train_data <- log(train_data)
log_test_data <- log(test_data)

```

```{r}

arima_log_model <- auto.arima(log_train_data)

summary(arima_log_model)

```

```{r}

# Forecast for test period
forecast_values <- forecast(arima_log_model, h = length(log_test_data))

# Plot actual vs forecast
autoplot(forecast_values) + 
  autolayer(log_test_data, series="Actual") + 
  ggtitle("ARIMA Forecast vs Actual Data") + 
  theme_minimal()

```

```{r}

# Ljung-Box Test for residual autocorrelation
Box.test(arima_log_model$residuals, lag = 20, type = "Ljung-Box")

```

```{r}

# Normality Tests
shapiro_test <- shapiro.test(arima_log_model$residuals)
print(shapiro_test)

jarque_bera_test <- jarque.bera.test(arima_log_model$residuals)
print(jarque_bera_test)

checkresiduals(arima_log_model)

```

# 12.1 SARIMA Model

```{r}

# Fit SARIMA model
sarima_log_model <- auto.arima(log_train_data, seasonal = TRUE)

# Summary of SARIMA model
summary(sarima_log_model)

```

```{r}

# Forecast for test period
forecast_values <- forecast(sarima_log_model, h = length(log_test_data))

# Plot actual vs forecast
autoplot(forecast_values) + 
  autolayer(log_test_data, series="Actual") + 
  ggtitle("ARIMA Forecast vs Actual Data") + 
  theme_minimal()

```

```{r}

# Ljung-Box Test for residual autocorrelation
Box.test(sarima_log_model$residuals, lag = 20, type = "Ljung-Box")

```

```{r}

# Normality Tests
shapiro_test <- shapiro.test(sarima_log_model$residuals)
print(shapiro_test)

jarque_bera_test <- jarque.bera.test(sarima_log_model$residuals)
print(jarque_bera_test)

checkresiduals(sarima_log_model)

```
# 13.0 AUTO.ARIMA Model BC

```{r}

# Apply Box-Cox Transformation (automatic lambda selection)
lambda <- BoxCox.lambda(train_data)  # Find optimal lambda
bc_train_data <- BoxCox(train_data, lambda)

lambda <- BoxCox.lambda(test_data)  # Find optimal lambda
bc_test_data <- BoxCox(test_data, lambda) 

```

```{r}

arima_bc_model <- auto.arima(bc_train_data)

summary(arima_bc_model)

```

```{r}

# Forecast for test period
forecast_values <- forecast(arima_bc_model, h = length(bc_test_data))

# Plot actual vs forecast
autoplot(forecast_values) + 
  autolayer(bc_test_data, series="Actual") + 
  ggtitle("ARIMA Forecast vs Actual Data") + 
  theme_minimal()

```

```{r}

# Ljung-Box Test for residual autocorrelation
Box.test(arima_bc_model$residuals, lag = 20, type = "Ljung-Box")

```

```{r}

# Normality Tests
shapiro_test <- shapiro.test(arima_bc_model$residuals)
print(shapiro_test)

jarque_bera_test <- jarque.bera.test(arima_bc_model$residuals)
print(jarque_bera_test)

checkresiduals(arima_bc_model)

```



# 13.1 SARIMA Model BC

```{r}

# Fit SARIMA model
sarima_bc_model <- auto.arima(bc_train_data, seasonal = TRUE)

# Summary of SARIMA model
summary(sarima_bc_model)

```

```{r}

# Forecast for test period
forecast_values <- forecast(sarima_bc_model, h = length(bc_test_data))

# Plot actual vs forecast
autoplot(forecast_values) + 
  autolayer(bc_test_data, series="Actual") + 
  ggtitle("ARIMA Forecast vs Actual Data") + 
  theme_minimal()

```

```{r}

# Ljung-Box Test for residual autocorrelation
Box.test(sarima_bc_model$residuals, lag = 20, type = "Ljung-Box")

```

```{r}

# Normality Tests
shapiro_test <- shapiro.test(sarima_bc_model$residuals)
print(shapiro_test)

jarque_bera_test <- jarque.bera.test(sarima_bc_model$residuals)
print(jarque_bera_test)

checkresiduals(sarima_bc_model)

```
