---
title: "ARIMA/SARIMA"
author: "Hritik"
date: "2025-03-03"
output: html_document
---

```{r}

# Load required libraries
library(tidyverse)
library(lubridate)
library(forecast)
library(tseries)
library(readr)
library(dplyr)
library(ggplot2)
library(lmtest) 

# Load dataset
setwd("/Users/hritikjhaveri/Desktop/UChicago/Q2/Time_Series_Analysis_Forcasting/Final_Project")
getwd()

data <- read.csv("London_AQI_Daily.csv")

# Convert 'Date' column to datetime format
data$Date <- as.Date(data$Date)

df <- data.frame(Date = data$Date, USAQI = data$USAQI)

# View first few rows
head(df)

# Check data types
str(df)

# Check for missing values
sum(is.na(df))

# Summary stats for AQI
summary(df$USAQI)

```

# 2. Visualize Time Series

```{r}

# Line plot
ggplot(df, aes(x = Date, y = USAQI)) +
  geom_line(color = "blue") +
  labs(title = "AQI in London from 2014-2024", x = "Year", y = "AQI in London") +
  theme_minimal()

```

# 3. Loading data into TS object

```{r}

# Decompose Time Series
ts_data <- ts(df$USAQI)

str(ts_data)

```

# 4. ADF Test

```{r}

# Augmented Dickey-Fuller (ADF) test
adf.test(ts_data)

```
# 5. KPSS Test

```{r}

# Kwiatkowski-Phillips-Schmidt-Shin (KPSS) test
kpss.test(ts_data)

```

# 6. Autocorrelation Function

```{r}

# ACF plots
acf(ts_data)

```

# 7. Partial Autocorrelation Function

```{r}

# PACF plots
pacf(ts_data)

```
# 8. Splitting data into training and testing sets

```{r}

# Find train indices
train_start_index <- which(df$Date == as.Date("2014-01-01"))
train_end_index <- which(df$Date == as.Date("2024-12-31"))

# Extract train window using indices
train_data <- window(ts_data, start = train_start_index, end = train_end_index)

# Find test indices
test_start_index <- which(df$Date == as.Date("2025-01-01"))
test_end_index <- which(df$Date == as.Date("2025-02-28"))

# Extract test window using indices
test_data <- window(ts_data, start = test_start_index, end = test_end_index)

```


# 9. ARIMA Model

```{r}

# Fit ARIMA model
arima_model <- auto.arima(train_data, seasonal = FALSE)

# Summary of ARIMA model
summary(arima_model)

```

```{r}

# Forecast for test period
forecast_values <- forecast(arima_model, h = length(test_data))

# Plot actual vs forecast
autoplot(forecast_values) + 
  autolayer(test_data, series="Actual") + 
  ggtitle("ARIMA Forecast vs Actual Data") + 
  theme_minimal()

```

# 10. SARIMA Model

```{r}

# Fit SARIMA model
sarima_model <- auto.arima(train_data, seasonal = TRUE)

# Summary of SARIMA model
summary(sarima_model)
```

```{r}
# Forecast for test period
forecast_values <- forecast(sarima_model, h = length(test_data))

# Plot actual vs forecast
autoplot(forecast_values) + 
  autolayer(test_data, series="Actual") + 
  ggtitle("SARIMA Forecast vs Actual Data") + 
  theme_minimal()

```


```{r}

checkresiduals(arima_model)
checkresiduals(sarima_model)

```

## Conclusion

ARIMA model is better since it has lower AIC and BIC values compared to SARIMA model. The residuals of ARIMA model are normally distributed and do not show any patterns, indicating that the model is a good fit for the data. The forecasted values are also closer to the actual values in the ARIMA model compared to the SARIMA model. Therefore, we can conclude that the ARIMA model is more suitable for forecasting AQI in London.