---
title: "TS ETS model final project"
author: "Anusha Bhat"
date: "2025-03-06"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

# Data Loading

```{r}
library(readxl)
library(forecast)
```

```{r}
# Read in data
aqi = read_excel("London_AQI_Daily.xlsx")

head(aqi)
```

```{r}
# Fix formatting 
aqi$Date = as.Date(aqi$Date, format = "%Y-%m-%d")
aqi$USAQI = as.integer(aqi$USAQI)
```

# EDA 

```{r}
plot(y = aqi$USAQI, x = aqi$Date, main = "Time Series of London AQI", type = "l",
     ylab = "AQI", xlab = "Date")
```

Looks seasonal and fairly consistent over time.


```{r}
# Convert df to ts 
aqi_ts = ts(aqi$USAQI, frequency = 365, start = c(2013, 1, 2))
```

```{r}
acf(aqi_ts, main = "ACF of AQI")
```

```{r}
pacf(aqi_ts, main = "PACF of AQI")
```

Lags have autocorrelation based on ACF but pacf shows limited autocorrelation when removing effect of short term lags.

```{r}
adf.test(aqi_ts)
```

```{r}
kpss.test(aqi_ts)
```

ADF test shows it's stationary but KPSS shows it's not so it may have stochastic and deterministic trends. Can difference it by one to confirm stationarity.

```{r}
new_aqi = diff(aqi_ts)
```

```{r}
adf.test(new_aqi)
```

```{r}
kpss.test(new_aqi)
```
After differencing, both tests confirm stationarity.


```{r}
acf(new_aqi, main = "ACF After Differencing")
```

```{r}
pacf(new_aqi, main= "PACF After Differencing")
```

There is still autocorrelation after differencing.

```{r}
decomp = stl(new_aqi, s.window = "periodic")
plot(decomp, main = "Seasonal Decomposition")
```

There seems to be a decreasing but stable trend over time with seasonality. 

```{r}
# Make train and test split 
train = aqi[aqi$Date < as.Date("2025-01-01"), ]
test = aqi[aqi$Date >= as.Date("2025-01-01"), ]

# Check
head(train)
tail(train)
head(test)
tail(test)
```

# Train/Test Split: Test = 2025 data

```{r}
# Convert to ts --> want 2: regular and differenced by 1
train_ts = ts(train$USAQI, frequency = 365, start = c(2013, 1, 2))
test_ts = ts(test$USAQI, frequency = 365, start = c(2025, 1, 1))

diff_train_ts = diff(train_ts)
diff_test_ts = diff(test_ts)
```


# ETS Model With No Differencing


```{r}
ets_mod = ets(train_ts)
summary(ets_mod)
```
```{r}
for1 = forecast(ets_mod, h = 59)
plot(for1, main = "Forecast for 2025", ylab = "AQI", xlab = "Date")
```

```{r}
checkresiduals(ets_mod)
```

Residuals have autocorrelation


```{r}
for1_vals = for1$mean
resids = test_ts - for1_vals

plot(resids, main = "Residuals for Test Data", ylab = "Residuals", xlab = "Time")
```

```{r}
rmse1 <- sqrt(mean(resids^2))

rmse1
```
RMSE is 61.733


```{r}
accuracy(ets_mod)
```

## ETS Model With Differencing

```{r}
ets_mod_diff = ets(diff_train_ts)
summary(ets_mod_diff)
```

```{r}
for2 = forecast(ets_mod_diff, h = 59)
plot(for2, main = "Forecast for 2025 (Differenced)", ylab = "AQI", xlab = "Date")
```

```{r}
checkresiduals(ets_mod_diff)
```

Residuals have autocorrelation


```{r}
for2_vals = for2$mean
resids2 = test_ts - for2_vals

plot(resids2, main = "Residuals for Test Data (Differenced)",
     ylab = "Residuals", xlab = "Time")
```

```{r}
rmse2 <- sqrt(mean(resids2^2))

rmse2
```
RMSE is 53.17281, lower than without differencing


```{r}
accuracy(ets_mod_diff)
```
RMSE on train is lower than RMSE without differencing.

Differenced model performs better. 












