---
title: "Cleaned Modeling"
author: "Anusha Bhat"
date: "2025-03-12"
output: html_document
---


```{r}
library(readxl)
library(tseries)
library(forecast)
library(hts)
library(zoo)
```

# Data loading (EDA in other file)

```{r}
# Read in data
aqi = read_excel("London_AQI_Daily.xlsx")

head(aqi)
```

```{r}
dailyaqi = aqi
dailyaqi$Date <- as.Date(dailyaqi$Date, format="%Y-%m-%d")
aqi_ts <- ts(dailyaqi$USAQI, frequency = 365, start = c(2013, 1))
```


```{r}
# split the data into train and test 
train_data <- subset(dailyaqi, Date < as.Date("2025-1-01"))
test_data  <- subset(dailyaqi, Date >= as.Date("2025-1-01"))

# create time series 
aqi_train_ts <- ts(train_data$USAQI, frequency = 365, start = c(2013, 1))
aqi_test_ts  <- ts(test_data$USAQI, frequency = 365, start = c(2025, 1))
```

```{r}
#### Apply differencing ####

# first-order differencing 
diff_train <- diff(aqi_train_ts, differences = 1)

# differenced box-cox 
boxcox_train <- BoxCox(aqi_train_ts, lambda = (BoxCox.lambda(aqi_train_ts)))

diff_boxcox_train <- diff(boxcox_train, differences = 1)

# seasonal differencing 
seasonal_train <- diff(aqi_train_ts, lag = 365)
```

Doing the following three transformations to account for outliers and to make the time series stationary:
-- differencing by 1
-- boxcox + differencing by 1
-- seasonal differencing (yearly)

```{r}
plot(stl(aqi_train_ts, s.window = "periodic"))
```

```{r}
plot(stl(diff_train, s.window = "periodic"))
```
```{r}
plot(stl(diff_boxcox_train, s.window = "periodic"))
```

```{r}
plot(stl(boxcox_train, s.window = "periodic"))
```
```{r}
plot(stl(seasonal_train, s.window = "periodic"))
```

The differenced and differenced box_cox had no non-zero trend so we will only use the original, boxcox, and seasonal for our models. 



Next, we will make a simple exponential smoothing model and a holt winters model for all 3 of these transformations.


### ETS Model 

#### No Difference 
```{r}
ets_mod = ets(aqi_train_ts)
summary(ets_mod)
```


```{r}
checkresiduals(ets_mod)
```

```{r}
# Forecast 
for1 = forecast(ets_mod, h = 59)
plot(for1, main = "Forecast for 2025 (No Transformation)", ylab = "AQI", xlab = "Date")
```


```{r}
plot(fitted(ets_mod), residuals(ets_mod), 
     main="Residuals vs Fitted Values", 
     xlab="Fitted Values", 
     ylab="Residuals",
     pch = 16)
abline(h=0, col="red")
```
```{r}
for1_vals = forecast(ets_mod)$mean
resids1 = aqi_test_ts - for1_vals
rmse1 = sqrt(mean(resids1^2))

cat("Test RMSE:", rmse1)
```
The residuals are autocorrelated which is expected. 

#### BoxCox


```{r}
bc_ets_mod = ets(boxcox_train)
summary(bc_ets_mod)
```


```{r}
checkresiduals(bc_ets_mod)
```


```{r}
# Forecast 
bc_for = forecast(bc_ets_mod, h = 59)
plot(bc_for, main = "Forecast for 2025 (Box-Cox)", ylab = "AQI", xlab = "Date")
```


```{r}
plot(fitted(bc_ets_mod), residuals(bc_ets_mod), 
     main="Residuals vs Fitted Values", 
     xlab="Fitted Values", 
     ylab="Residuals",
     pch = 16)
abline(h=0, col="red")
```

```{r}
reverse_boxcox = function(predicted, original, lambda){
  if (lambda == 0){
    exp(predicted)
  }
  else {
    (lambda * predicted + 1)^(1 / lambda)
  }
}
```


```{r}
# reverse differencing and box-cox 
lambda_value = BoxCox.lambda(aqi_train_ts)

forecast_values_bc = reverse_boxcox(bc_for$mean, aqi_train_ts, lambda_value)
```

```{r}
bc_resids = aqi_test_ts - forecast_values_bc
bc_rmse = sqrt(mean(bc_resids^2))

cat("Test RMSE:", bc_rmse)
```
The residuals are still autocorrelated



#### Differenced Seasonal 


```{r}
ds_ets_mod = ets(seasonal_train)
summary(ds_ets_mod)
```


```{r}
checkresiduals(ds_ets_mod)
```

```{r}
# Forecast 
ds_for = forecast(ds_ets_mod, h = 59)
plot(ds_for, main = "Forecast for 2025 (Differenced Seasonal)", ylab = "AQI", 
     xlab = "Date", xlim = c(2024, 2025))
```


```{r}
plot(fitted(ds_ets_mod), residuals(ds_ets_mod), 
     main="Residuals vs Fitted Values", 
     xlab="Fitted Values", 
     ylab="Residuals",
     pch = 16)
abline(h=0, col="red")
```



```{r}
original = aqi_train_ts
forecast_values_season = (original[(length(original) - 365 + 1):(length(original) - 365 + 59)]) + ds_for$mean
ds_resids = aqi_test_ts - forecast_values_season
ds_rmse = sqrt(mean(ds_resids^2))

cat("Test RMSE:", ds_rmse)
```

The residuals are autocorrelated. This performed better than no transformation but not as good as boxcox which was the best. 



### Holt Winters 


#### No Differencing


```{r}
hw_mod = HoltWinters(aqi_train_ts, seasonal = "additive")
summary(hw_mod)
```


```{r}
hw_for = forecast(hw_mod, h = 59)
plot(hw_for, main = "Forecast for 2025 HW (No Transformation)")
```


```{r}
plot(hw_mod$fitted[,1], residuals(hw_mod), 
     main="Residuals vs Fitted Values", 
     xlab="Fitted Values", 
     ylab="Residuals",
     pch = 16)
abline(h=0, col="red")
```


```{r}
checkresiduals(hw_mod)
```

```{r}
forecast_values_hw =  hw_for$mean
hw_res = aqi_test_ts - forecast_values_hw

hw_rmse = sqrt(mean(hw_res^2))

cat("Test RMSE:", hw_rmse)
```
This performed better than the general ets. Residuals still autocorrelated.


#### BoxCox

```{r}
bc_hw_mod = HoltWinters(boxcox_train, seasonal = "additive")
summary(bc_hw_mod)
```


```{r}
bc_hw_for = forecast(bc_hw_mod, h = 59)
plot(bc_hw_for, main = "Forecast for 2025 HW (BoxCox)")
```


```{r}
plot(bc_hw_mod$fitted[,1], residuals(bc_hw_mod), 
     main="Residuals vs Fitted Values", 
     xlab="Fitted Values", 
     ylab="Residuals",
     pch = 16)
abline(h=0, col="red")
```


```{r}
checkresiduals(bc_hw_mod)
```

```{r}
# reverse differencing and box-cox 
lambda_value = BoxCox.lambda(aqi_train_ts)

forecast_values_hw_bc = reverse_boxcox(bc_hw_for$mean, aqi_test_ts, lambda_value)
```



```{r}
bc_hw_res = aqi_test_ts - forecast_values_hw_bc

bc_hw_rmse = sqrt(mean(bc_hw_res^2))

cat("Test RMSE:", bc_hw_rmse)
```


#### Seasonal Differencing 


```{r}
ds_hw_mod = HoltWinters(seasonal_train, seasonal = "additive")
summary(ds_hw_mod)
```


```{r}
ds_hw_for = forecast(ds_hw_mod, h = 59)
plot(ds_hw_for, main = "Forecast for 2025 HW (Differenced Seasonal)")
```


```{r}
plot(ds_hw_mod$fitted[,1], residuals(ds_hw_mod), 
     main="Residuals vs Fitted Values", 
     xlab="Fitted Values", 
     ylab="Residuals",
     pch = 16)
abline(h=0, col="red")
```


```{r}
checkresiduals(ds_hw_mod)
```

```{r}
original = aqi_train_ts
forecast_values_hw_season = (original[(length(original) - 365 + 1):(length(original) - 365 + 59)]) + ds_hw_for$mean
ds_resids = aqi_test_ts - forecast_values_hw_season

ds_hw_rmse = sqrt(mean(ds_resids^2))

cat("Test RMSE:", ds_hw_rmse)
```



### Model comparisons

```{r}
#ETS
ets_rmse_table <- data.frame(
  "No Transformation" = rmse1,
  "Boxcox" = bc_rmse,
  "Seasonal" = ds_rmse
)
rownames(ets_rmse_table) <- c("Test Accuracy")

ets_rmse_table
```

### Model comparisons

```{r}
#HW
hw_rmse_table <- data.frame(
  "No Transformation" = hw_rmse,
  "Boxcox" = bc_hw_rmse,
  "Seasonal" = ds_hw_rmse
)
rownames(hw_rmse_table) <- c("Test Accuracy")

hw_rmse_table
```

Holt-winters made seasonal and boxcox perform worse but no transformation performed better. Boxcox was the best performing overall.



# Plots: 


```{r}
par(mfrow = c(1, 3))
acf(residuals(ets_mod), main = "ETS No Transformation")
acf(residuals(bc_ets_mod), main = "ETS Box-Cox")
acf(residuals(ds_ets_mod), main = "ETS Seasonally Differenced")
```
```{r}
plot(stl(aqi_train_ts, s.window = "periodic"), 
     main = "STL Decomposition (No Transformation)")
plot(stl(boxcox_train, s.window = "periodic"), 
     main = "STL Decomposition (Box-Cox)")
plot(stl(seasonal_train, s.window = "periodic"), 
     main = "STL Decomposition (Seasonally Differenced)")
plot(stl(diff_train, s.window = "periodic"), 
     main = "STL Decomposition (1st Order Differenced)")
plot(stl(diff_boxcox_train, s.window = "periodic"), 
     main = "STL Decomposition (Seasonally Differenced)")
```
```{r}
#getting last 60 values from train
train_last_60 <- tail(aqi_train_ts, 60)

# Plot forecast with test data (on original scale)
plot(forecast_values_bc, main = "ETS Forecast (Box-Cox Transformation)", ylim = c(0, 150 ), xlim = c(2024.85,2025.18), ylab="AQI")
lines(aqi_test_ts, col = "red", lwd = 2)
lines(train_last_60, col = "blue", lwd = 2)
legend("topright", legend = c("Forecast", "Test"), col = c("black", "red"), lty = 1)

```

```{r}
plot(bc_for, xlim = c(2024.8, 2025))
```

























































