---
title: "Cleaned Modeling"
author: "Anusha Bhat"
date: "2025-03-12"
output: html_document
---


```{r}
library(readxl)
library(tseries)
library(forecast)
library(hts)
library(zoo)
```

# Data loading (EDA in other file)

```{r}
# Read in data
aqi = read_excel("London_AQI_Daily.xlsx")

head(aqi)
```

```{r}
dailyaqi = aqi
dailyaqi$Date <- as.Date(dailyaqi$Date, format="%Y-%m-%d")
aqi_ts <- ts(dailyaqi$USAQI, frequency = 365, start = c(2013, 1))
```


```{r}
# split the data into train and test 
train_data <- subset(dailyaqi, Date < as.Date("2025-1-01"))
test_data  <- subset(dailyaqi, Date >= as.Date("2025-1-01"))

# create time series 
aqi_train_ts <- ts(train_data$USAQI, frequency = 365, start = c(2013, 1))
aqi_test_ts  <- ts(test_data$USAQI, frequency = 365, start = c(2025, 1))
```

```{r}
#### Apply differencing ####

# first-order differencing 
diff_train <- diff(aqi_train_ts, differences = 1)

# differenced box-cox 
boxcox_train <- BoxCox(aqi_train_ts, lambda = (BoxCox.lambda(aqi_train_ts)))

diff_boxcox_train <- diff(boxcox_train, differences = 1)

# seasonal differencing 
seasonal_train <- diff(aqi_train_ts, lag = 365)
```

Doing the following three transformations to account for outliers and to make the time series stationary:
-- differencing by 1
-- boxcox + differencing by 1
-- seasonal differencing (yearly)


Next, we will make a simple exponential smoothing model and a holt winters model for all 3 of these transformations.


### ETS Model on Differenced Data 

#### Differenced Regular 
```{r}
diff_ets_mod = ets(diff_train)
summary(diff_ets_mod)
```


```{r}
checkresiduals(diff_ets_mod)
```

```{r}
# Forecast 
diff_for = forecast(diff_ets_mod, h = 59)
plot(diff_for, main = "Forecast for 2025 (Differenced)", ylab = "AQI", xlab = "Date")
```


```{r}
plot(fitted(diff_ets_mod), residuals(diff_ets_mod), 
     main="Residuals vs Fitted Values", 
     xlab="Fitted Values", 
     ylab="Residuals",
     pch = 16)
abline(h=0, col="red")
```

```{r}
forecast_values_diff = cumsum(c(tail(train_data$USAQI, 1), diff_for$mean))[-1] 
diff_resids = aqi_test_ts - forecast_values_diff
diff_rmse = sqrt(mean(diff_resids^2))

cat("Test RMSE:", diff_rmse)
```
The residuals are autocorrelated, so let's incorporate AR into the model

```{r}
diff_ar = auto.arima(diff_train)
summary(diff_ar)
```
Differenced TS has AR(5)

```{r}
checkresiduals(diff_ar)
```


```{r}
Box.test(residuals(diff_ets_mod) + residuals(diff_ar), lag = 10)
```
```{r}
diff_for_ar = forecast(diff_ar, h = 59)
diff_preds = forecast_values_diff + diff_for_ar$mean

diff_cets_rmse = sqrt(mean((aqi_test_ts - diff_preds)^2))
cat("Test RMSE:", diff_cets_rmse)
```
Residuals are still autocorrelated and the test RMSE went up.


#### Differenced BoxCox


```{r}
dbc_ets_mod = ets(diff_boxcox_train)
summary(dbc_ets_mod)
```


```{r}
checkresiduals(dbc_ets_mod)
```


```{r}
# Forecast 
dbc_for = forecast(dbc_ets_mod, h = 59)
plot(diff_for, main = "Forecast for 2025 (Differenced BC)", ylab = "AQI", xlab = "Date")
```


```{r}
plot(fitted(dbc_ets_mod), residuals(dbc_ets_mod), 
     main="Residuals vs Fitted Values", 
     xlab="Fitted Values", 
     ylab="Residuals",
     pch = 16)
abline(h=0, col="red")
```
```{r}
# reverse differencing and box-cox 
lambda_value = BoxCox.lambda(aqi_train_ts)z

forecast_values_bc = cumsum(c(tail(train_data$USAQI, 1), dbc_for$mean))[-1]
# forecast_values_real <- inv_boxcox(forecast_values_bc, lambda = lambda_value)

# have to shift values to ensure positive values since some values in (forecast_values_bc * lambda_value + 1) are negative, causing the exponentiation to fail. 
forecast_values_bc = ((forecast_values_bc - min(forecast_values_bc) + 1) * lambda_value + 1)^(1/lambda_value)
```

```{r}
dbc_resids = aqi_test_ts - forecast_values_bc
dbc_rmse = sqrt(mean(dbc_resids^2))

cat("Test RMSE:", dbc_rmse)
```
The residuals are autocorrelated, so let's incorporate AR into the model


```{r}
dbc_ar = auto.arima(boxcox_train)
summary(dbc_ar)
```

Differenced TS has AR(2)

```{r}
checkresiduals(dbc_ar)
```


```{r}
Box.test(residuals(dbc_ets_mod) + residuals(dbc_ar), lag = 10)
```


```{r}
dbc_for_ar = forecast(dbc_ar, h = 59)
dbc_preds = forecast_values_bc + dbc_for_ar$mean

dbc_cets_rmse = sqrt(mean((aqi_test_ts - dbc_preds)^2))
cat("Test RMSE:", dbc_cets_rmse)
```
Residuals are still autocorrelated but they decreased a little with using the AR.



#### Differenced Seasonal 


```{r}
ds_ets_mod = ets(seasonal_train)
summary(ds_ets_mod)
```


```{r}
checkresiduals(ds_ets_mod)
```

```{r}
# Forecast 
ds_for = forecast(ds_ets_mod, h = 59)
plot(ds_for, main = "Forecast for 2025 (Differenced Seasonal)", ylab = "AQI", 
     xlab = "Date", xlim = c(2024, 2025))
```


```{r}
plot(fitted(ds_ets_mod), residuals(ds_ets_mod), 
     main="Residuals vs Fitted Values", 
     xlab="Fitted Values", 
     ylab="Residuals",
     pch = 16)
abline(h=0, col="red")
```



```{r}
#forecast_values_season = cumsum(c(tail(train_data$USAQI, 365), ds_for$mean))[-(1:365)]
original = aqi_train_ts
forecast_values_season = (original[(length(original) - 365 + 1):(length(original) - 365 + 59)]) + ds_for$mean
ds_resids = aqi_test_ts - forecast_values_season
ds_rmse = sqrt(mean(ds_resids^2))

cat("Test RMSE:", ds_rmse)
```

The residuals are autocorrelated, so let's incorporate AR into the model

```{r}
ds_ar = auto.arima(seasonal_train)
summary(ds_ar)
```

Seasonal TS has AR(5)

```{r}
checkresiduals(ds_ar)
```


```{r}
Box.test(residuals(ds_ets_mod) + residuals(ds_ar), lag = 10)
```


```{r}
ds_for_ar = forecast(ds_ar, h = 59)
ds_preds = forecast_values_season + ds_for_ar$mean

ds_cets_rmse = sqrt(mean((aqi_test_ts - ds_preds)^2))
cat("Test RMSE:", ds_cets_rmse)
```
Residuals are still autocorrelated but the RMSE increased a little with using the AR.



### Holt Winters 


#### Differenced Model 


```{r}
diff_hw_mod = HoltWinters(diff_train, seasonal = "additive")
summary(diff_hw_mod)
```


```{r}
diff_hw_for = forecast(diff_hw_mod, h = 59)
plot(diff_hw_for, main = "Forecast for 2025 HW (Differenced)")
```


```{r}
plot(diff_hw_mod$fitted[,1], residuals(diff_hw_mod), 
     main="Residuals vs Fitted Values", 
     xlab="Fitted Values", 
     ylab="Residuals",
     pch = 16)
abline(h=0, col="red")
```


```{r}
checkresiduals(diff_hw_mod)
```

```{r}
forecast_values_hw_diff =  cumsum(c(tail(train_data$USAQI, 1), 
                                    diff_hw_for$mean))[-1] 
diff_hw_res = aqi_test_ts - forecast_values_hw_diff

diff_hw_rmse = sqrt(mean(diff_hw_res^2))

cat("Test RMSE:", diff_hw_rmse)
```
Incorporating AR(5):

```{r}
Box.test(residuals(diff_hw_mod) + residuals(diff_ar), lag = 10)
```

```{r}
diff_preds2 = forecast_values_hw_diff + diff_for_ar$mean

diff_chw_rmse = sqrt(mean((aqi_test_ts - diff_preds2)^2))
cat("Test RMSE:", diff_chw_rmse)
```
Residuals are still autocorrelated and the test had higher RMSE. The 


#### Differenced BoxCox

```{r}
dbc_hw_mod = HoltWinters(diff_boxcox_train, seasonal = "additive")
summary(dbc_hw_mod)
```


```{r}
dbc_hw_for = forecast(dbc_hw_mod, h = 59)
plot(diff_hw_for, main = "Forecast for 2025 HW (Differenced BC)")
```


```{r}
plot(dbc_hw_mod$fitted[,1], residuals(dbc_hw_mod), 
     main="Residuals vs Fitted Values", 
     xlab="Fitted Values", 
     ylab="Residuals",
     pch = 16)
abline(h=0, col="red")
```


```{r}
checkresiduals(dbc_hw_mod)
```
```{r}
# reverse differencing and box-cox 
lambda_value = BoxCox.lambda(aqi_train_ts)

forecast_values_hw_bc = cumsum(c(tail(train_data$USAQI, 1), dbc_hw_for$mean))[-1]
# forecast_values_real <- inv_boxcox(forecast_values_bc, lambda = lambda_value)

# have to shift values to ensure positive values since some values in (forecast_values_bc * lambda_value + 1) are negative, causing the exponentiation to fail. 
forecast_values_hw_bc = ((forecast_values_hw_bc - min(forecast_values_hw_bc) + 1) * lambda_value + 1)^(1/lambda_value)
```



```{r}
dbc_hw_res = aqi_test_ts - forecast_values_hw_bc

dbc_hw_rmse = sqrt(mean(dbc_hw_res^2))

cat("Test RMSE:", dbc_hw_rmse)
```
Incorporating AR(2):

```{r}
Box.test(residuals(dbc_hw_mod) + residuals(dbc_ar), lag = 10)
```

```{r}
dbc_preds2 = forecast_values_hw_diff + diff_for_ar$mean

dbc_chw_rmse = sqrt(mean((aqi_test_ts - dbc_preds2)^2))
cat("Test RMSE:", dbc_chw_rmse)
```

Residuals are still autocorrelated but the test had much lower RMSE.



#### Seasonal Differencing 


```{r}
ds_hw_mod = HoltWinters(seasonal_train, seasonal = "additive")
summary(ds_hw_mod)
```


```{r}
ds_hw_for = forecast(ds_hw_mod, h = 59)
plot(diff_hw_for, main = "Forecast for 2025 HW (Differenced Seasonal)")
```


```{r}
plot(ds_hw_mod$fitted[,1], residuals(ds_hw_mod), 
     main="Residuals vs Fitted Values", 
     xlab="Fitted Values", 
     ylab="Residuals",
     pch = 16)
abline(h=0, col="red")
```


```{r}
checkresiduals(ds_hw_mod)
```

```{r}
original = aqi_train_ts
forecast_values_hw_season = (original[(length(original) - 365 + 1):(length(original) - 365 + 59)]) + ds_hw_for$mean
ds_resids = aqi_test_ts - forecast_values_hw_season

ds_hw_rmse = sqrt(mean(ds_resids^2))

cat("Test RMSE:", ds_hw_rmse)
```

Incorporating AR(2):

```{r}
Box.test(residuals(ds_hw_mod) + residuals(ds_ar), lag = 10)
```

```{r}
ds_preds2 = forecast_values_hw_season + diff_for_ar$mean

ds_chw_rmse = sqrt(mean((aqi_test_ts - ds_preds2)^2))
cat("Test RMSE:", ds_chw_rmse)
```

Residuals are still autocorrelated but the test had similar RMSE.


### Model comparisons

```{r}
#ETS
ets_rmse_table <- data.frame(
  "Differenced" = diff_rmse,
  "Differenced + ARIMA" = diff_cets_rmse,
  "Differenced Boxcox" = dbc_rmse,
  "Differenced Boxcox + ARIMA" = dbc_cets_rmse,
  "Seasonal" = ds_rmse,
  "Seasonal + ARIMA" = ds_cets_rmse
)
rownames(ets_rmse_table) <- c("Test Accuracy")

ets_rmse_table
```

### Model comparisons

```{r}
#HW
hw_rmse_table <- data.frame(
  "Differenced" = diff_hw_rmse,
  "Differenced + ARIMA" = diff_chw_rmse,
  "Differenced Boxcox" = dbc_hw_rmse,
  "Differenced Boxcox + ARIMA" = dbc_chw_rmse,
  "Seasonal" = ds_hw_rmse,
  "Seasonal + ARIMA" = ds_chw_rmse
)
rownames(hw_rmse_table) <- c("Test Accuracy")

hw_rmse_table
```

Holt-winters made seasonal and differenced perform worse but boxcox differenced performed better. For both, differenced was the best performing, though it had lower RMSE for the simple ets model.





































































